Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis

### 2018-06-17 article/Old versions/MPS/PCADSC_article.tex
angle-plot updated!

\subsection*{The angle plot}
The angle plot compares both loadings and eigenvalues at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e. loading and eigenvalue pairs) are most similar and most different across the two datasets. Let $\lambda_{\max} = \max\{ \lambda_{1}^x, \lambda_{1}^y \}$ be the largest eigenvalue for the two datasets. Then the empirical correlation matrix for the $\mathbf{X}$ dataset, $S_x$, has the following orthogonal decomposition in the coordinate system of the $\mathbf{Y}$ dataset
\begin{equation*}
S_x = \sum_{j=1}^d \lambda_{j}^x \eta_{j}^x ({\eta_{j}^x})^\top
= \lambda_{\max} \sum_{k=1}^d
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \eta_{j}^y ({\eta_{j}^y}^\top \eta_{k}^x) \Bigg)
{\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \eta_{j}^y ({\eta_{j}^y}^\top \eta_{k}^x) \Bigg)}^\top,
\end{equation*}
and we have a similar decomposition of $S_y$ in the coordinate system of the $\mathbf{X}$ dataset. We propose to visualize these two decompositions in a $d \times d$ grid display. In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell. The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ counterclockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal. To facilitate the following description we will refer to the arrows drawn counterclockwise as the blue arrows, and the arrows drawn clockwise as the red arrows. The lengths $\mu_{jk}$ and $\nu_{jk}$, and the angle $\theta_{jk}$, are given by
\begin{align*}
\mu_{jk} &= \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \lvert {\eta_{k}^x}^\top \eta_{j}^y \rvert, &
\nu_{jk} &= \sqrt{\frac{\lambda_{j}^y}{\lambda_{\max}}} \lvert {\eta_{j}^y}^\top \eta_{k}^x \rvert, &
\theta_{jk} &= \acos(\lvert {\eta_{k}^x}^\top \eta_{j}^y \rvert).
\end{align*}
For two $d$-dimensional, unit length vectors $a$ and $b$, it holds that $ a^\top b = \langle a, b \rangle = \tilde{c}(a,b)$, where $\tilde{c}$ denotes the empirical correlation function. Thus, in the angle plot, we are essentially looking at the absolute values of correlations between loadings that have been scaled according to their variance contributions. The absolute value of the projection ${\eta_{k}^x}^\top \eta_{j}^y$ is inserted due to the indeterminacy of the direction of loading vectors. This indeterminacy implies that the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_x$ and $S_y$ can be visualized in a joint plot by dividing the angles by two and using counterclockwise and clockwise shifts from the diagonal. Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length. Figure \ref{plot.simAngle} presents two examples of angle plots based on simulated data.

@@

\hl{For dataset $\mathbf{A}$, we find very short off-diagonal arrows, suggesting that the two PCA decompositions agree on the relative placement of the information in  the data among the 6 PCs.  For dataset $\mathbf{B}$, on the other hand, we find very little agreement in all components, which should be expected, as the two groups in the dataset correspond to different data structures.}

In the \emph{angle plot}, the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal component of the $\mathbf{A}$ dataset in the coordinate system of the $\mathbf{B}$ dataset. Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components of the $\mathbf{B}$ dataset in the coordinate system of the $\mathbf{A}$ dataset. \hl{If the structures of the two datasets are identical, then we will have coinciding blue and red arrows along the diagonal in the grid display, and nothing else as arrows in the off-diagonal cells would have zero length. Differences in the eigenvalues are visualized as differences in the lengths of the blue and the red arrows, also in the diagonal. And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells, as well as arrows of non vanishing length in the off-diagonal cells.}

>>>>

\subsection*{The angle plot}
The angle plot compares both loadings and eigenvalues at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e. loading and eigenvalue pairs) are most similar and most different across the two datasets. Let $\lambda_{\max} = \max\{ \lambda_{1}^x, \lambda_{1}^y \}$ be the largest eigenvalue for the two datasets. Then the empirical correlation matrix for the $\mathbf{X}$ dataset, $S_x$, has the following orthogonal decomposition in the coordinate system of the $\mathbf{Y}$ dataset
\begin{equation*}
S_x = \sum_{j=1}^d \lambda_{j}^x \eta_{j}^x ({\eta_{j}^x})^\top
= \lambda_{\max} \sum_{k=1}^d
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \eta_{j}^y ({\eta_{j}^y}^\top \eta_{k}^x) \Bigg)
{\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \eta_{j}^y ({\eta_{j}^y}^\top \eta_{k}^x) \Bigg)}^\top,
\end{equation*}
and we have a similar decomposition of $S_y$ in the coordinate system of the $\mathbf{X}$ dataset. We propose to visualize these two decompositions in a $d \times d$ grid display. In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell. The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ counterclockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal. To facilitate the following description we will refer to the arrows drawn counterclockwise as the blue arrows, and the arrows drawn clockwise as the red arrows. The lengths $\mu_{jk}$ and $\nu_{jk}$, and the angle $\theta_{jk}$, are given by
\begin{align*}
\mu_{jk} &= \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \lvert {\eta_{k}^x}^\top \eta_{j}^y \rvert, &
\nu_{jk} &= \sqrt{\frac{\lambda_{j}^y}{\lambda_{\max}}} \lvert {\eta_{j}^y}^\top \eta_{k}^x \rvert, &
\theta_{jk} &= \acos(\lvert {\eta_{k}^x}^\top \eta_{j}^y \rvert).
\end{align*}
For two $d$-dimensional, unit length vectors $a$ and $b$, it holds that $ a^\top b = \langle a, b \rangle = \tilde{c}(a,b)$, where $\tilde{c}$ denotes the empirical correlation function. Thus, in the angle plot, we are essentially looking at the absolute values of correlations between loadings that have been scaled according to their variance contributions. The absolute value of the projection ${\eta_{k}^x}^\top \eta_{j}^y$ is inserted due to the indeterminacy of the direction of loading vectors. This indeterminacy implies that the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_x$ and $S_y$ can be visualized in a joint plot by dividing the angles by two and using counterclockwise and clockwise shifts from the diagonal. Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.

In the \emph{angle plot}, the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal component of the $\mathbf{A}$ dataset in the coordinate system of the $\mathbf{B}$ dataset. Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components of the $\mathbf{B}$ dataset in the coordinate system of the $\mathbf{A}$ dataset. \hl{If we have coinciding blue and red arrows along the diagonal in the grid display, and arrows of vanishing lengths in the off-diagonal cells, then the structures of the two datasets are identical. In contrast, differences in the eigenvalues are visualized as differences in the lengths of the blue and the red arrows, also in the diagonal. And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells, as well as arrows of non vanishing length in the off-diagonal cells.}

\hl{However, there is a generic situation where the angles between the arrows lack discriminative power, namely when two or more neighbouring eigenvalues are identical or close. For instance, multiple eigenvalues close to 1 arise if more than one of the variables are uncorrelated, or close to uncorrelated, of all the other variables. In this situation only the corresponding eigenspace is well estimates, but the rotation of the associated eigenvectors will be subject to sampling variability. To investigate if we are in such a situation of identical data structures with neighbouring eigenvalues being close, where comparably long arrows may occur in the corresponding blocks of off-diagonal cells, we have implemented a permutation test similar as done for the CE-plot. For each random reallocation we recompute the angles $\theta_{jk}$. The resampled angles may both be used for quantitative tests, where large angles are critical for the diagonal cells and small angles are critical for the off-diagonal cells, as well as for the visualisation of the sampled angle regions. In the visualisation we propose to present the entire sampled region of angles for each cell, but grey shaded according to percentiles such that more critical angles are lighter grey. In our implementation we do not present the p-values for the quantitative tests, but the p-values are available in the R-object to support the visualization if necessary. However, it is important to remember that there are $d \times d$ p-values, which means that attention should be given to a possible multiple testing issue.}

@@

\hl{Figure \ref{plot.simAngle} presents two examples of angle plots based on simulated data. For dataset $\mathbf{A}$, we find very short off-diagonal arrows except for the off-diagonal cells corresponding to the 3'rd and 4'th eigenvalues. However, we see that the arrows in cell (3,4) and (4,3) lie well within the shaded angle regions (with associated p-values 0.9119 and 0.9103). This indicates that we are in a situation with identical data structure, and with 3'rd and 4'th eigenvalues being close (this corresponds to the two independent variables in \emph{model 1}). For dataset $\mathbf{B}$, on the other hand, we find very little agreement in all components, which should be expected, as the two groups in the dataset correspond to different data structures.}
